---
title: "Parallel chained equations"
author: "Marc Lane"
date: "17/08/2026"
output: html_document
---

This document showcases the algorithms developed for my thesis "Chained Equations for Incomplete Climate Networks", supervised by Brian O'Sullivan. This work builds on two imputation algorithms introduced in Brian's paper (https://doi.org/10.1002/joc.8513) Elastic-net Chained Equations (ENCE) and Multiple Imputation by Chained Equations with Direct Use of Regularised Regression (MICE DURR).

ENCE and MICE DURR were originally designed for filling missing data in spatiotemporal rainfall records collected by Met Éireann's monitoring network across the Republic of Ireland. ENCE and MICE DURR are built using the chained equations method, which is an iterative imputation procedure that fits a separate regression model for every incomplete variable in sequence. My thesis introduced a novel modification to chained equations called synchronous chained equations. Synchronous chained equations fits each model in a cycle at the same time, which makes the algorithm parallel.

This file compares 3 implementations of ENCE and MICE DURR: the original serial versions written entirely in R, the synchronous versions also in R, and the C implementation with synchronous updates parallelised across threads using OpenMP. This repository uses simulated daily data (data/simulated_daily_rainfall.csv), which simulates 1 year of daily rainfall in Ireland across 200 stations. The user can edit data/simulate_rainfall.Rmd to create a different dataset.


## Setup

```{r setup, message=FALSE}

# Compiling C code
so_path <- "src/ENCE_MICE_DURR_impute.so" # find .so file if you already compiled it
# if C code hasn't been compiled yet
if (!file.exists(so_path)) {
  cat("Compiling C source code... \n")
  build_log <- suppressWarnings(
  system2("bash", c("-c", shQuote("cd src && bash build.sh")), stdout = TRUE, stderr = TRUE)
)
  cat(build_log, sep = "\n")
  if (!file.exists(so_path)) stop("Compiling failed: see compiler output above.")
}

source("R/idw.R")
source("R/load_library.R")

# Set the number of cores on your machine for synchronous updates
# I usually set the number of cores manually here (n_cores <- 6L), since parallel::detectCores(logical = FALSE) doesn't reliably work on WSL/Linux
n_cores <- parallel::detectCores(logical = FALSE) # See README note on github.
if (is.na(n_cores)) {
  n_cores <- 2L  # assume there's 2 cores
  warning("Could not detect physical core count using parallel::detectCores(logical = FALSE), defaulting to 2 threads. Manually set n_cores to the number of cores on your machine.")
}
cat("Using n_cores =", n_cores, " for parallel synchronous updates\n")


# Load simulated data
rain <- read.csv("data/simulated_daily_rainfall.csv")
rain <- rain[, c("stno", "east", "north", "t", "y")]

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Mask a fraction of the observed cells to validate against
set.seed(541)
observed_idx <- which(!is.na(rain$y))
masked_idx   <- sort(sample(observed_idx, length(observed_idx) * 0.25))

df_miss <- rain
df_miss$y[masked_idx] <- NA           # hide masked
```

## Run Imputations

Run the original R serial implementation of ENCE:

```{r run-ENCE-R-serial, message=FALSE, results='hide'}
source("R/ENCE_serial.R")
time_ence_serial <- system.time({
  imp_ence_serial <- ENCE_serial(df_miss, response = "y", max_cycles = 30,
                     init_method = "idw",
                     x_name = "east", y_name = "north", t_name = "t",
                     C = 15000,
                     transformation = sqrt,
                     reverse_transformation = function(x) x^2,
                     truth = rain$y, masked_idx = masked_idx,
                     lambda_options = c(0.2, 0.15, 0.1, 0.05),
                     alpha_options  = c(0.1, 0.35, 0.65, 0.9))
})
```

Run the synchronous R version of ENCE:

```{r run-ENCE-R-synchronous, message=FALSE, results='hide'}
source("R/ENCE_synchronous.R")
time_ence_sync <- system.time({
  imp_ence_sync <- ENCE_sync(df_miss, response = "y", max_cycles = 30,
                     init_method = "idw",
                     x_name = "east", y_name = "north", t_name = "t",
                     C = 15000,
                     transformation = sqrt,
                     reverse_transformation = function(x) x^2,
                     truth = rain$y, masked_idx = masked_idx,
                     lambda_options = c(0.2, 0.15, 0.1, 0.05),
                     alpha_options  = c(0.1, 0.35, 0.65, 0.9),
                     n_workers = n_cores)
})
```

Run the C implementation (with synchronous updates) of ENCE:

```{r run-ENCE-C, message=FALSE, results='hide'}
source("R/ENCE_C.R")
time_ence_C <- system.time({
  imp_ence_C <- ENCE_C(df_miss, response = "y", max_cycles = 30,
                     init_method = "idw",
                     x_name = "east", y_name = "north", t_name = "t",
                     C = 15000,
                     transformation = sqrt,
                     reverse_transformation = function(x) x^2,
                     truth = rain$y, masked_idx = masked_idx,
                     lambda_options = c(0.2, 0.15, 0.1, 0.05),
                     alpha_options  = c(0.1, 0.35, 0.65, 0.9),
                     nthreads = n_cores)
})
```

Run the original R serial implementation of MICE DURR (with m=1 for manageable runtime):

```{r run-MICE_DURR_serial, message=FALSE, results='hide'}
source("R/MICE_DURR_serial.R")
time_mice_serial <- system.time({
  imp_mice_serial <- MICE_DURR_serial(df_miss, response = "y", m = 1,
                                 max_impute_cycles = 5, hyp_cycles = 2, 
                                 init_method = "idw",
                                 max_EM_cycles = 5,
                                 nclusters = 1,
                                 lambda_options = c(0.2, 0.15, 0.1, 0.05),
                                 alpha_options  = c(0.1, 0.35, 0.65, 0.9),
                                 outfile = NULL,
                                 transformation = sqrt,
                                 reverse_transformation = {function(x) x^2})
})
```

Run the synchronous R version of MICE DURR:

```{r run-MICE_DURR_synchronous, message=FALSE, results='hide'}
source("R/MICE_DURR_synchronous.R")
time_mice_sync <- system.time({
  imp_mice_sync <- MICE_DURR_sync(df_miss, response = "y", m = 1,
                                 max_impute_cycles = 5, hyp_cycles = 2, 
                                 init_method = "idw",
                                 max_EM_cycles = 5,
                                 nclusters = 1,
                                 n_workers = n_cores,
                                 lambda_options = c(0.2, 0.15, 0.1, 0.05),
                                 alpha_options  = c(0.1, 0.35, 0.65, 0.9),
                                 outfile = NULL,
                                 transformation = sqrt,
                                 reverse_transformation = {function(x) x^2})
})
```

Run the C implementation (with synchronous updates) of MICE DURR:

```{r run-MICE_DURR_C, message=FALSE, results='hide'}
source("R/MICE_DURR_C.R")
time_mice_C <- system.time({
  imp_mice_C <- MICE_DURR_C(df_miss, response = "y", m = 1,
                                 max_impute_cycles = 5, hyp_cycles = 2, 
                                 init_method = "idw",
                                 max_EM_cycles = 5,
                                 nclusters = 1,
                                 nthreads = n_cores,
                                 lambda_options = c(0.2, 0.15, 0.1, 0.05),
                                 alpha_options  = c(0.1, 0.35, 0.65, 0.9),
                                 outfile = NULL,
                                 transformation = sqrt,
                                 reverse_transformation = {function(x) x^2})
})
```


## Results
```{r metrics}
y_true <- rain$y[masked_idx]
y_ence_serial <- imp_ence_serial$y[masked_idx]
y_ence_sync   <- imp_ence_sync$y[masked_idx]
y_ence_C      <- imp_ence_C$y[masked_idx]
y_mice_serial <- imp_mice_serial$y[masked_idx]
y_mice_sync   <- imp_mice_sync$y[masked_idx]
y_mice_C      <- imp_mice_C$y[masked_idx]

# Helper function for printing metrics
report_results <- function(y_hat, y_true, time, speedups = NULL) {
  t_self <- time[["elapsed"]]

  cat("==========================================================\n")
  cat("Time: ", round(t_self, 2), "seconds\n")

  for (nm in names(speedups)) {
    cat(paste0("Speedup, ", nm, ":"),
        round(speedups[[nm]][["elapsed"]] / t_self, 1), "x\n")
  }

  rmse <- sqrt(mean((y_hat - y_true)^2))

  cat("==========================================================\n")
  cat("Validation Metrics:\n")
  cat("RMSE:",   rmse, "\n")
  cat("Bias %:", sum(y_hat - y_true) / sum(y_true) * 100, "\n")
  cat("R2:",     1 - sum((y_hat - y_true)^2) / sum((y_true - mean(y_true))^2), "\n")
  cat("SRMSE:",  rmse / mean(y_true), "\n\n")

  invisible(NULL)
}
```
## ENCE results:

```{r ENCE-serial-results, results='hold'}
cat("ENCE R Serial:\n")
report_results(y_ence_serial, y_true, time_ence_serial)
```

```{r ENCE-sync-results, results='hold'}
cat("ENCE R Synchronous:\n")
report_results(y_ence_sync, y_true, time_ence_sync, speedups = list("R serial -> R parallel" = time_ence_serial))
```

```{r ENCE-C-results, results='hold'}
cat("ENCE C implementation:\n")
report_results(y_ence_C, y_true, time_ence_C,
               speedups = list("R serial -> C implementation" = time_ence_serial, "R sync -> C implementation"   = time_ence_sync))
```

## MICE DURR Results

```{r MICE-DURR-serial-results, results='hold'}
cat("MICE DURR R Serial:\n")
report_results(y_mice_serial, y_true, time_mice_serial)
```

```{r MICE-DURR-sync-results, results='hold'}
cat("MICE DURR R Synchronous:\n")
report_results(y_mice_sync, y_true, time_mice_sync,
               speedups = list("R serial -> R parallel" = time_mice_serial))
```

```{r MICE-DURR-C-results, results='hold'}
cat("MICE DURR C Synchronous:\n")
report_results(y_mice_C, y_true, time_mice_C,
               speedups = list("R serial -> C implementation" = time_mice_serial, "R sync -> C implementation"   = time_mice_sync))
```
